---
title: "Differential gene expression analysis (DESeq)"
author: "Dmytro Kryvokhyzha"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/git/bioinformatic-pipelines/DESeq2')
```

# Libraries

```{r libraries, message=FALSE, echo=FALSE}
library(DESeq2)
library(vsn)
library(pheatmap)
library(RColorBrewer)
library(IHW)
library(ggplot2)
```

# Load the data

```{r rna-data}
tissues <- c ('pancreas', 'muscle') # list of tissues
for (t in tissues){
  d <- read.table(paste('data/RNA-Seq_mmquant_', t, '.csv', sep=""),
                  header=TRUE, row.names='Gene')
  assign(paste(t, 'D', sep=""), d)
  message(t, ": samples: ", dim(d)[2], ", genes: ", dim(d)[1])
}
rm(d,t)
dannot <- read.csv('data/RNA-Seq_annot.csv', header = T)
```

# Functions

I use [IHW p-value correction](https://www.nature.com/articles/nmeth.3885).

```{r}
runDEseq <- function(tissue, d, annot){
  # annot <- dannot
  # tissue <- 'pancreas'
  # d <- get(paste('d', tissue, sep=""))
  samplesID <- as.character(annot$Sample_id[(annot$Tissue==tissue) & 
                                             annot$Sample_id %in% colnames(d)])

  annot <- annot[annot$Sample_id %in% samplesID,]
  d <- d[,samplesID]
  
  species <- as.character(annot$Species[annot$Sample_id %in% samplesID])
  sex <- as.character(annot$Sex[annot$Sample_id %in% samplesID])
  age <- as.character(annot$Age[annot$Sample_id %in% samplesID])
  breed <- as.character(annot$Breed[annot$Sample_id %in% samplesID])
  colData <- cbind(samplesID, species, sex, age, breed)
  
  dds <- DESeqDataSetFromMatrix(countData = d[,samplesID],
                                colData = colData,
                                design= ~ sex + age + species)
  dds <- DESeq(dds)
  res <- results(dds, contrast=c('species', 'wolf', 'dog'), filterFun=ihw)
  #res <- lfcShrink(dds, coef="species_wolf_vs_dog", type="ashr")
  resOrdered <- res[order(res$padj),]
  return(list(resOrdered, dds))
}

plotPCAfun <- function(res, name){
  vstRes <- vst(res, blind=T)
  plotPCA(vstRes, intgroup=c("species")) + #xlim(-40, 40) + ylim(-40, 40) +
          geom_text(label=colnames(res), size=3, hjust=0.5, vjust=-1)
}
```

# Explore the data

Load unfiltered data to DEseq objects:

```{r, message=FALSE}
for (t in tissues){
  d <- get(paste(t, 'D', sep=""))
  Res0 <- runDEseq(t, d, dannot)
  assign(paste(t, 'Res0', sep=""), Res0)
}
rm(d,t)
```

## PCA

```{r, message=FALSE}
for (t in tissues){
  d <- get(paste(t, 'Res0', sep=""))
  p <- plotPCAfun(d[[2]], t)
  print(p + ggtitle(t))
}
rm(d,t)
```

## Library size

**Multi-mapped reads included.**

To asses the expression of duplicated genes, I use the
[multi-map groups (MMGs)](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-015-0734-x#Sec16)
method implemented in [mmquant](https://bitbucket.org/mzytnicki/multi-mapping-counter/src/master/)

```{r  message=FALSE, fig.width=10, fig.height=5}
libsize <- c()
for (t in tissues){
  d <- get(paste(t, 'D', sep=""))
  lsd <- colSums(d)
  libsize <- append(libsize, lsd)
}
annot <- dannot
annot <- annot[which(annot$Sample_id %in% names(libsize)),]
rownames(annot) <- annot$Sample_id
annot <- annot[names(libsize),]
barplot(libsize/1000000, names.arg=names(libsize), border = F,
        col= as.character(annot$Col), ylab = "Library size (million reads)", las =2,
        cex.names=0.6)
```

**Multi-mapped reads excluded.**

```{r  message=FALSE, fig.width=10, fig.height=5}
libsizeNoDupl <- c()
for (t in tissues){
  dd <- get(paste(t, 'D', sep=""))
  dd <- dd[-c(grep("_", rownames(dd))),]
  lsdd <- colSums(dd)
  libsizeNoDupl <- append(libsizeNoDupl, lsdd)
}
barplot(libsizeNoDupl/1000000, names.arg=names(libsizeNoDupl), border = F, ylim = c(0,80), 
        col= as.character(annot$Col), ylab = "Library size (million reads)", las =2,
        cex.names=0.6)
```

# Run the analyses

## Remove outliers

Based on the PCA and prior knowledge (e.g. potentially degradeted samples),
I removed the outliers:

```{r}
outlierSamples <- c('Wolf_P15', 'Wolf_M6', 'Wolf_M7')
for (t in tissues){
  d <- get(paste(t, 'D', sep=""))
  Dfilter <- d[,!(colnames(d) %in% outlierSamples)]
  assign(paste(t, 'Dfilter', sep=""), Dfilter)
}
rm(d,Dfilter,t)
```

## Differential expression analysis

```{r, message=FALSE}
for (t in tissues){
  d <- get(paste(t, 'Dfilter', sep=""))
  Res <- runDEseq(t, d, dannot)
  assign(paste(t, 'Res', sep=""), Res)
}
rm(d,t)
```

Summary of the results:

```{r}
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  message(t)
  summary(d[[1]])
}
rm(d,t)
```

Note, the defaul **alpha is 0.1**, here is the number of DE genes for **0.05**:

```{r}
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  s <- sum(d[[1]]$padj < 0.05, na.rm=TRUE)
  message(paste(t, s, sep = ": "))
}
rm(d,s)
```

Number of DE MMG (percentage):

```{r}
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  s <- sum(d[[1]][grep("_", rownames(d[[1]])),]$padj < 0.05, na.rm=TRUE)
  total <- length(d[[1]][grep("_", rownames(d[[1]])),]$padj)
  perc <- round(s/total, 2)
  message(paste(t, s, sep = ": "), paste(" (", perc, ")", sep = ""))
}
rm(d,s)
```

Cook’s distances to further check if there are any outliers:

```{r}
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  boxplot(log10(assays(d[[2]])[["cooks"]]), range=0, las=2, main=t)
}
rm(d)
```

Dispersion:

```{r}
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  plotDispEsts(d[[2]], main=t)
}
rm(d)
```

Scatter plot of fold changes vs normalized counts.

```{r}
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  plotMA(d[[1]], ylim=c(-2,2), main=t)
}
```

Look at specific MMG/gene (e.g. AMY2B):

```{r echo=FALSE}
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  gene <- rownames(d[[1]])[grep("ENSCAFG00000028653", rownames(d[[1]]))]
  plotCounts(d[[2]], gene=gene,
           intgroup="species", main = paste("AMY2B", t, sep = " "))
}
```

```{r echo=FALSE}
message("Tssue\tlogFC\tpadj")
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  amyFC <- d[[1]][grep("ENSCAFG00000028653", rownames(d[[1]])),]$log2FoldChange
  amyPval <- d[[1]][grep("ENSCAFG00000028653", rownames(d[[1]])),]$padj
  message(t, "\t", amyFC, "\t", amyPval)
}
```

Data transformations for visualization

```{r}
for (t in tissues){
  d <- get(paste(t, 'Res', sep=""))
  vst <- vst(d[[2]], blind=T)
  assign(paste(t, 'Vst', sep=""), vst)
}
```

Heatmap of the count matrix

```{r}
for (t in tissues){
  res <- get(paste(t, 'Res', sep=""))
  vst <- get(paste(t, 'Vst', sep=""))
  select <- order(rowMeans(counts(res[[2]], normalized=TRUE)),
                decreasing=TRUE)[1:50]
  d <- as.data.frame(colData(res[[2]])[,c('sex', 'age')])
  pheatmap(assay(vst)[select,], cluster_rows=FALSE, show_rownames=FALSE,
           cluster_cols=FALSE, annotation_col=d, main=t)
}
```

Sample-to-sample distances:

```{r}
for (t in tissues){
  vst <- get(paste(t, 'Vst', sep=""))
  sampleDists <- dist(t(assay(vst)))
  sampleDistMatrix <- as.matrix(sampleDists)
  rownames(sampleDistMatrix) <- paste(vst$samplesID, vst$age, vst$sex, sep="-")
  colnames(sampleDistMatrix) <- NULL
  colors <- colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)
  pheatmap(sampleDistMatrix,
           clustering_distance_rows=sampleDists,
           clustering_distance_cols=sampleDists,
           col=colors, main=t)
}
```

### PCA without outliers

```{r}
for (t in tissues){
  vst <- get(paste(t, 'Vst', sep=""))
  p <- plotPCA(vst, intgroup=c("species")) +
          geom_text(label=colnames(vst), size=3, hjust=0.5, vjust=-1)
  print(p + ggtitle(t))
}
```

Write the result to files:

```{r eval = FALSE}
for (t in tissues){
  res <- get(paste(t, 'Res', sep=""))[[1]]
  resSig <- res[which(res$padj < 0.05),]
  genesSig <- rownames(resSig[grep("_", rownames(resSig), invert =T),])
  mmgSign <- rownames(resSig[grep("_", rownames(resSig)),])
  nMMG <- length(mmgSign)
  genesMMG <- unique(unlist(strsplit(mmgSign, "_")))
  lenGenesMMG <- length(unlist(strsplit(mmgSign, "_")))
  lenUniqGenesMMG <- length(genesMMG)
  message(paste(t, ":", nMMG, "MMG;", lenGenesMMG,  'genes;', lenUniqGenesMMG, "unique;",  sep = " "))
  write.csv(as.data.frame(res), 
          file=paste("results/tables/DESeq2_mmquant_", t, ".csv", sep = ""))
  write.table(as.data.frame(genesSig), row.names = F, quote = F, col.names = F,
          file=paste("scratch/GO/DESeq2_mmquant_", t, "_sign.genes", sep = ""))
  write.table(as.data.frame(genesMMG), row.names = F, quote = F, col.names = F,
          file=paste("scratch/GO/DESeq2_mmquant_", t, "_sign.mmg", sep = ""))
}
```
